Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-03-11 code/eda-code/eda.qmd
Part 3 Updates
I started the statistical analysis and created 6 simple linear models. I added to the methods section in the manuscript and created a second version of the processed code based on the eda-code.

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr) # for basic piping functions
library(skimr) # for summarizing data
library(ggplot2) # for visualizing data
library(gtsummary) # for creating tables
```

Load the data.

```{r}
#Path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reprodutive age. 
```{r}
# population size standarization for total counts per birth control method
mydata <- mydata %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```

@@

>>>>

```{r}
#load necessary packages
library(here) #for data loading/saving
library(dplyr) # for basic piping functions
library(skimr) # for summarizing data
library(ggplot2) # for visualizing data
library(gtsummary) # for creating tables
```

# Load the data

```{r}
#path to data
data_location <- here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

@@

After reviewing the raw data set again, we found the `wra` variable that accounts for the number of women of reproductive age (15-49) in each country. Once this variable was added to the original processed data set, we could use it to standardize the total counts for each birth control method by population size of women of reprodutive age. 
```{r}
# population size standardization for total counts per birth control method
mydata <- mydata %>% mutate(larcster_standard = total_larcster / wra,
                            sarc_standard = total_sarc / wra,
                            trad_standard = total_trad / wra, 
                            none_standard = total_none / wra)
```

@@

# Save data 
Lastly, we will save this data with the new standardized total usage variables for each method of birth control as the second version of our processed data with only the variables necessary for the statistical analysis and modeling portion of the project. This new dataset is saved in the same location as the first processed dataset. 
```{r}
# select variables necessary for statistical analysis/modeling 
processeddata2 <- mydata %>% select(country, wra, region, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad,  pct_upreg_nouse, larcster_usage, sarc_usage, trad_usage, none_usage, larcster_standard, sarc_standard, trad_standard, none_standard)

# save it to the same folder with the first version of the processed data
save_data_location <- here("data","processed-data","processeddata2.rds")
saveRDS(processeddata2, file = save_data_location)
```

### 2024-03-11 code/processing-code/processing-file.qmd
Part 3 Updates
I started the statistical analysis and created 6 simple linear models. I added to the methods section in the manuscript and created a second version of the processed code based on the eda-code.

code for low, medium, and high use per country using case_when()

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

Fortunately, the rest of the variables required for this analysis are complete. Each of the four variables describing the percentage of unintended pregnancies based on birth control type also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100)

@@

We should address the missing data in the general `pct_upreg` here to avoid introducing bias because 18.18% of this variable is missing, and it is essential for answering the research question. After reviewing the codebook, the researchers provided an external source for this variable, which we have included in the references. The numbers in this source do not match what is included in the data set. After exploring the methodology for the Adding It Up study, we found their original data that matches what is recorded in the dataset. When we tried to create our own percentage of unintended pregnancy variable, we found that each birth control method adds up to account for 100% of unintended pregnancies. Since there are 24 observations missing in what should be the total unintended pregnancy variable, it is likely that the variables with the percentages of unintended pregnancies by birth control method were imputed or estimated. Missingness in the collection of data could also explain why the long-acting reversible contraceptives and sterilization method were combined into one category. Based on this discovery, we modified the question that we want to answer with the data to use variables that are complete and reliable: `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse`. These variables will allow us to answer the question of is there a significant difference in the percentage of unplanned pregnancies between three contraceptive methods among all countries compared to no usage?

>>>>

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, wra, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, ster_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, ster_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534, ster_2534, sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, ster_3549, sarc_3549, tradmethods_3549, nomethod_3549)

@@

Fortunately, the rest of the variables required for this analysis are complete including the `wra` variable with population size that we will need to standardize population counts later in the exploratory data analysis process. Each of the four variables describing the percentage of unintended pregnancies based on birth control type also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100)

@@

We should address the missing data in the general `pct_upreg` here to avoid introducing bias because 18.18% of this variable is missing, and it is essential for answering the research question. After reviewing the codebook, the researchers provided an external source for this variable, which we have included in the references. The numbers in this source do not match what is included in the data set. After exploring the methodology for the Adding It Up study, we found their original data that matches what is recorded in the dataset. When we tried to create our own percentage of unintended pregnancy variable, we found that each birth control method adds up to account for 100% of unintended pregnancies. Since there are 24 observations missing in what should be the total unintended pregnancy variable, it is likely that the variables with the percentages of unintended pregnancies by birth control method were imputed or estimated. Missingness in the collection of data could also explain why the long-acting reversible contraceptives and sterilization method were combined into one category. Based on this discovery, we modified the question that we want to answer with the data to use variables that are complete and reliable: `pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`, and `pct_upreg_nouse`. These variables will allow us to answer the question of is there a significant difference in the percentage of unplanned pregnancies between three contraceptive methods among all countries compared to no usage.

### 2024-03-11 products/manuscript/Manuscript.qmd
Part 2 Error
Added sessionInfo and looked for reasons why others would get an error when rendering the manuscript.
Checking for errors
Found the version of Quarto we are using and checked that the renv package is updated.
Part 3 Updates
I started the statistical analysis and created 6 simple linear models. I added to the methods section in the manuscript and created a second version of the processed code based on the eda-code.

# Research Questions
1) Is there a significant difference in the percentage of unplanned pregnancies between contraceptive methods among all countries? Does this difference vary by region?
The main outcome variables will be total percentage of unintended pregnancies per birth control method (`pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`). 
The control variable will be total percentage of unintended pregnancies among women not using birth control (`pct_upreg_none`)
The predictor variables will include the total number of people women using each type of birth control standardized by population size, which will be utilized to create a factor variable showing low, medium, and high usage for each method per country. 
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

@@

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across age groups (20-24) and (35-49) to investigate the effectiveness of each contraceptive method.Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group. Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study. The table provided above under `Summarize the data` shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists, but regulations differ. Chile, Colombia, and Ecuador legally recognize access, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision to victims of sexual and domestic violence (Hevia, 2012).

@@

1. Is there a significant difference in the percentage of unplanned pregnancies between three  contraceptive methods among all countries compared to no usage?

2. How does marriage status affect percentage of usage across three different types of birth control per region? (back-up question)

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:
Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

@@

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

>>>>

# Research Questions
1) Is there a significant difference in the percentage of unplanned pregnancies between contraceptive methods among all countries? Does this difference vary by region?
The main outcome variables will be total percentage of unintended pregnancies per birth control method (`pct_upreg_larcster`, `pct_upreg_sarc`, `pct_upreg_trad`). 
The control variable will be total percentage of unintended pregnancies among women not using birth control (`pct_upreg_none`)
The predictor variables will include the total number of people women using each type of birth control standardized by population size, which will be utilized to create a factor variable showing low, medium, and high usage for each method per country. 
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be combined across age groups to create a total usage variable for each method. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

@@

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across all age groups (15-19, 20-24, 25-35, and 35-49) to investigate the effectiveness of each contraceptive method. Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group. Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study. The table provided above under `Summarize the data` shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists, but regulations differ. Chile, Colombia, and Ecuador legally recognize access, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision to victims of sexual and domestic violence (Hevia, 2012).

@@

1. Is there a significant difference in the percentage of unplanned pregnancies between three contraceptive methods and no usage between regions of the world?

@@

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
Reproducing this project requires R, RStudio, and Microsoft word. The README files in each folder contain descriptions of the documents located in that folder to help guide users through the reproducible processes. Begin with the processing-file.qmd in the code > processing-code folder. This file loads the original dataset from the Guttmacher Institute's Adding It Up project, loads the codebook for the dataset, and glimpses at the data. We chose 30 variables to clean to find the percentage of unintended pregnancies based on each type of contraceptive methods: long acting reversible and sterilization methods, short acting reversible methods, traditional methods, and no contraceptive usage. After the missing data was explored, we chose to use the percentage of unintended pregnancies variables related to each method of birth control because the total variable had unexplained missingness. The processed data is saved in the data > processed-data folder as processeddata.rds.

The next step is exploratory data analysis. The eda.qmd file in the code > eda-code folder contains all the code for this process. We produce a couple summary tables before we begin a simple exploration of trends in percentage of unintended pregnancies by country in our first set of histograms. The second set of histograms explore contraceptive usage methods by country. These graphs revealed outliers, so we standardized by population size for each country. The next sets of plots are line graphs showing total usage of contraceptive methods to percentage of unintended pregnancies associated with that method. All of these figures are saved in the Supplementary-Material.qmd file in the products > manuscript > supplement folder. The last two sets of plots contain line graphs of the relationship between the two variables stratified by region and violin plots displaying the relationship between percentage of unintended pregnancies and the categorical usage variable created during processing, which are included in this manuscript. The exploratory data analysis process led to another version of the processed data with standardized total usage variables saved as processeddata2.rds in the data > procsssed-data folder.

Step three is the statistical analayis and model fitting portion found in the statistical-analysis.qmd file in the code > analysis-code folder.

The Manuscript.qmd file for this document can be found in the products > manuscript folder.
